---
title: "Beta Regression"
subtitle: "Team Bee (Beta Regressionists) (Advisor: Dr. Seals)"
author: "Anaite Montes Bu, Travis Keep"
editor: source
format: 
  revealjs: 
    toc: true
    theme: dark
    self-contained: true
    css: custom.css
    html-math-method: katex
bibliography: references1.bib
csl: ieee.csl
execute:
  warning: false
  message: false
  error: false
---

## Introduction

-   Regression analysis is a statistical tool used to explore relationships between variables.

-   Beta Regression: When the dependent variable is a ratio or percentage.

## Shortfalls of Normal Regression

- Assumes dependent variable is normally distributed
- Assumes variance is constant throughout

Generally not true with ratios or percentages

## When dependent variable lies in (0, 1)

- True of ratios or percents such as test scores
- Variance typically less near the extremes e.g 0 or 1.

## Reading Skills data set

- Reading skills based on IQ and if student has dyslexia N=44
- Normal regression shows IQ alone not significant in predicting reading score
- Beta regression shows IQ alone is significant in predicting reading score
[@smithson2006]

## Beta distribution

The PDF of random variable with a beta distribution is as follows.

$$
f(y) = \begin{cases} 
      \frac{y^{\alpha-1}(1-y)^{\beta-1}}{B(\alpha,\beta)}, & 0 \le y \le 1 \\
      0, & \text{elsewhere}
\end{cases}
$$
Where $B(\alpha,\beta) = \int_0^1 y^{\alpha-1}(1-y)^{\beta-1} \ dy = \frac{\Gamma(\alpha) \Gamma(\beta)}{\Gamma(\alpha+\beta)}.$

$\alpha$ and $\beta$ are the shape variables where $\alpha > 0 \quad \beta > 0$.
[@wackerly2002]

## Beta Distribution Mean and Variance

$$
E[Y] = \mu = \frac{\alpha}{\alpha+\beta} \ \ \ \text{and} \ \ \ V[Y] = \sigma^2  = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}
$$
[@wackerly2002]

## Introduction of $\mu$ and $\phi$

For beta regression, it is useful to introduce the following

$$
\mu = \frac{\alpha}{\alpha+\beta} \\
\phi = \alpha + \beta
$$
$\mu$ is the mean of the beta regression while the higher the $\phi$ the less the variance or the less spread out the PDF function is. [@ferrari2004]

## Revised Beta Distribution

$$
f(y; \mu, \phi) = \frac{\Gamma(\phi)}{\Gamma(\mu \phi) \Gamma((1 - \mu)\phi)} y^{\mu\phi - 1}(1 - y)^{(1 - \mu)\phi - 1}, \quad 0 < y < 1
$$
Where:\
- **(μ)** is the mean, **(ϕ)** is the precision (inverse of the variance), **(Γ)** is the gamma function.

## Beta Distribution Variance

$$
\text{Var}(Y) = \frac{\mu(1 - \mu)}{1 + \phi}
$$
When $\mu$ is near the extremes, 0 or 1, variance drops. [@Ferrari2004]

## Extended Beta Regression

**Bias** **Correction/Reduction**
- **Type** **of** **Estimator:**

- ML (Maximum Likelihood): Standard method, useful but may yield biased estimates in certain conditions.[@Grun2012]
- BC (Bias-Corrected): Adjusts estimates to correct for bias, providing more reliable parameter values.
- BR (Bias-Reduced): Shrinks estimates towards a central value, which can improve predictive performance.

## **Bias** **Correction/Reduction**

```{r}
library(betareg)
library(readr)
suicide_dataset <- read_csv("master.csv")

max_suicide_rate <- max(suicide_dataset$`suicides/100k pop`, na.rm = TRUE)
suicide_dataset$suicide_rate <- ifelse(suicide_dataset$`suicides/100k pop` == 0, 0.001,
                                        ifelse(suicide_dataset$`suicides/100k pop` == max_suicide_rate, 0.999,
                                               suicide_dataset$`suicides/100k pop` / max_suicide_rate))

# Rename variables
names(suicide_dataset)[names(suicide_dataset) == "HDI for year"] <- "HDI_year"
names(suicide_dataset)[names(suicide_dataset) == "gdp_per_capita ($)"] <- "GDP_capita"

# Remove rows with NA values
suicide_dataset <- na.omit(suicide_dataset)

# Mean model
m1 <- suicide_rate ~ HDI_year + GDP_capita + sex + age + generation | HDI_year + GDP_capita

# Fit beta regression models
suicide_ml <- betareg(m1, data = suicide_dataset, type = "ML")
suicide_bc <- betareg(m1, data = suicide_dataset, type = "BC")
suicide_br <- betareg(m1, data = suicide_dataset, type = "BR")

summary(suicide_bc)
```


## Beta Regression Trees

-   This extension uses recursive partitioning to model data that might exhibit subgroup-specific relationships. 
-   It builds decision trees by splitting data into different subgroups based on the instability of model parameters across partitioning variables.



```{r}
library(partykit)
set.seed(123)

# Randomly sample 20% of the data
sample_indices <- sample(seq_len(nrow(suicide_dataset)), size = 0.2 * nrow(suicide_dataset))
suicide_sample <- suicide_dataset[sample_indices, ]

# Fit the model on the sample
suicide_tree <- betatree(suicide_rate ~ HDI_year, 
                         data = suicide_sample, 
                         minsplit = 20,
                         control = betatree.control(maxdepth = 3))

# Plot the resulting tree
plot(suicide_tree)
```

## Diagnostic Plots
The package **betareg** allows users to perform both fixed and variable dispersion beta regression. The model is based on the beta distribution, using a parameterization with the mean and precision[@Zeileis2004].

```{r}
par(mfrow = c(2, 2))
plot(suicide_bc)
```



## Conclusion

-   Effective for proportion data, Ideal for modeling data bounded in the (0, 1) range.
-   Models both mean and precision, managing boundary cases and latent heterogeneity.
-   Bias correction and beta regression trees expand its capabilities.
-   The betareg package in R offers a powerful, flexible framework for analysts.

## References


